\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-16 paper/paper.tex
paper updates

At this point, we decided to take a different approach and use an approach similar to that used in Sprout~\cite{sprout}.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other. In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected. This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations. Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}. We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to 67\% higher.

In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate. For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate. In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.

We did notice that Minstrel occasionally has very poor performance on traces 1 seconds or shorter---sometimes as low as 2 Mbps. This is because it does not stop sending after four successive failures, but must wait until the end of the 100ms window for statistics to be compiled. Thus it wastes a lot of time sending at non-optimal rates, and a short trace only has a few windows. This behavior does not show up in long traces at all, however, as Minstrel has had time to both smooth out bad windows and collect more statistics (making it less likely to choose a bad rate).

bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly. The variations were small, though. On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps. SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps. The variation of results on other traces was similar.

The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bit-rates. SampleRate, however, spends too much time sampling low bit-rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates. In the specific case we examined, they were all probe packets. But sometimes Minstrel would choose 54 Mbps as the rate with the best throughput, which made little sense. A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54 Mbps. The probability of success would be 1, so then 54 Mbps would be chosen as the rate with the best throughput. This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54 Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate! This is because the EWMA did not account for the fact that the second block having sent hundreds more packets at 54 Mbps than the first. Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly. In essence, the probabilities were not declining fast enough after a large number of failures.

In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix the probability problem.

>>>>

At this point, we decided to take a different approach and use an trace-based method similar to that used in Sprout~\cite{sprout}.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other. In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected. This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations. Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}. We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, sometimes with throughput as much as 67\% higher.

In Figure~\ref{figure:2} we show the performance of constant bit-rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal constant rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bit-rate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit-rate selection is---the optimal bit-rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

Bit-rate selection algorithms strive to have a throughput near or greater than the optimal constant rate. For this \texttt{clear\_1.dat} trace, SampleRate clearly meets this goal, achieving an average throughput more than 1 Mbps higher than the optimal constant rate. In noisier situations, though, Minstrel tends to be closer to the optimal constant rate.

Bit-rate selection algorithms naturally show variation between runs on the same data because probe rates are chosen randomly. The variations were small, though. On ten runs on the \texttt{clear\_1.dat} trace, the average of Minstrel over 10 runs was 7.52 Mbps, with a maximum variation of +/- 0.12 Mbps. SampleRate was similarly stable, it averaged 5.284 Mbps over 10 runs, with a maximum variation of about +/- 0.3 Mbps. The variation of results on other traces was similar.

The histograms of which rates the algorithms decided to send at provided some useful insight into their behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bit-rates. SampleRate, however, spends too much time sampling low bit-rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

Our examination of the histograms showed us that Minstrel spends a lot of time sending at high rates. In the specific case we examined, they were all probe packets. But sometimes Minstrel would choose 54 Mbps as the rate with the best throughput, which made little sense. A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54 Mbps. The probability of success would be 1, so then 54 Mbps would be chosen as the rate with the best throughput. This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54 Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate! This is because the EWMA did not account for the fact that the second block having sent hundreds more packets at 54 Mbps than the first. Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly. In essence, the probabilities were not declining fast enough after a large number of failures.

Additionally, Minstrel uses a complex method of calculating when to send a sample packet that involves a heuristic to make sure that 10\% of packets were actually samples. This is because the mrr places sample packets second in the retry chain when the sample rate is slower than the rate with the best throughput. So even if 10\% of probes have a sample rate in the retry chain, a sample rate isn't actually sent 10\% of the time. We implemented this heuristic and it tended to cause Minstrel to send too many probe packets.

In the next section we will talk about Minproved, an improved version of Minstrel where we attempt to fix problems with EWMA and too-frequent probing
